Accessible, realistic genome simulation with selection using \stdpopsim

### 2025-07-06 analysis2_manuscript.tex
Response to reviewer factor of 2 comment, plus methods clarification.
Update analysis2_manuscript.tex
Here are my edits on the revision. I added some context on the vaquita in a couple places. I also switched out the word "benchmark" in a few places to make it seem less like we're claiming we did a proper benchmarking. This is in line with the spirit of the response to the last comment from the first reviewer.

\section*{Abstract}
    \label{abstract}
    Selection is a fundamental evolutionary force that shapes patterns of genetic variation across species.
    However, simulations incorporating realistic selection along heterogeneous genomes in complex demographic histories
    are challenging, limiting our ability to benchmark statistical methods aimed at detecting selection and to explore theoretical predictions.
    \stdpopsim is a community-maintained simulation library that already provides an extensive catalog of
    species-specific population genetic models.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of selection, including background selection, selective sweeps,
    and arbitrary distributions of fitness effects (DFE) acting on annotated subsets of the genome (for instance, exons).
    This extension maintains \stdpopsim's core principles of reproducibility and accessibility while adding support
    for species-specific genomic annotations and published DFE estimates.
    We demonstrate the utility of this framework by benchmarking methods for demographic inference,
    DFE estimation, and selective sweep detection across several species and scenarios.
    Our results demonstrate the robustness of demographic inference methods to selection on linked sites,
    reveal the sensitivity of DFE-inference methods to model assumptions,
    and show how genomic features, like recombination rate and functional sequence density, influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible, low-barrier framework.

@@

    As with other aspects of \stdpopsim, we intend the selection models to be as biologically accurate as possible.
    Thus, the \stdpopsim catalog provides \texttt{Annotation} objects based on species' publicly available functional genomic elements
    and \texttt{DFE} objects based on published DFE estimates (Figure \ref{fig:schematic}A).
    By using annotation objects, one can, for example, simulate populations where selected mutations (whose selection
    coefficients are drawn from a specified DFE) occur within annotation regions (e.g. exons), while all mutations
    occurring outside of these regions do not affect fitness.
    There are relatively few species with published DFE estimates in the literature.
    \stdpopsim currently has implemented DFEs for five species in the catalog:
    \textit{Arabidopsis thaliana} \citep{huber2018gene},
    \textit{Drosophila melanogaster} \citep{ragsdale2016triallelic,huber2017determining,zhen2021greater},
    humans \citep{huber2017determining,kim2017inference,kyriazis2023using,castellano2019compmarison,castellano2019compmarison,rodrigues2024shared},
    \textit{Mus musculus} \citep{booker2021selective},
    and the vaquita porpoise \textit{Phocoena sinus} \citep{robinson2022critically}.
    While the DFEs are from a limited number of species, these DFEs can be applied to other species
    if and when it is suitable,
    namely, when a species of interest lacks a pubished DFE in the catalog \revpoint{1}{8}.
    For instance, one could simulate a butterfly species with a gamma-distributed DFE originally
    estimated from humans, or a human population with a DFE estimated from \textit{Drosophila}.
    Indeed, the DFE labeled \texttt{HomSap/Mixed\_K23} is recommended by \citet{kyriazis2023using}
    for ``general'' use across mammals (lacking more specific information).
    Furthermore, the user can specify a custom DFE and provide their own annotations
    of functional elements to simulate selection in a species for which we do not yet have
    a published DFE included in the catalog.

@@

    To expand these observations to additional species and to highlight the ease of comparison enabled by \stdpopsim,
    we performed a similar benchmark using the vaquita porpoise \textit{Phocoena sinus}.
    We simulated 100 genomes under a two-epoch demographic model inferred for vaquita by \textcite{robinson2022critically},
    and a constant recombination rate across the genome (see \textbf{Methods}).
    In simulations with selection, we applied the DFE inferred by \textcite{robinson2022critically} to exons
    annotated from the vaquita genome assembly.
    This DFE implements a relationship between selection and dominance
    in which very deleterious mutations (with selection coefficient $s<-0.1$) are fully recessive
    (dominance coefficient $h=0$) and weakly deleterious mutations ($s\geq -0.001$) are nearly
    additive ($h=0.4$; see \textbf{Methods}).
    As in the human simulations, we see that the presence of selection
    decreases polymorphism in genomes simulated with selection, but that the effect
    is only readily apparent within exonic regions (Figure \ref{fig:vaquita-summary-stats}).
    For these simulations, the SFS-based method (\stairway) produces more accurate
    inference than the two methods based on the sequential Markovian coalescent
    (\smcpp and \msmc; see Figure \ref{fig:vaquita-demography}).
    Regardless of these differences, we see that background selection again does not
    appear to considerably influence the inference of any of the four methods for this species,
    at least for the parameterization used here,
    even though vaquita has a very different evolutionary history than humans.

@@

    Next, we benchmarked these three DFE-inference methods on simulations of the vaquita porpoise genome.
    Recall that these simulations apply a gamma-distributed selection coefficient to mutations in exons
    with a relationship between the selection and dominance coefficients (see \textbf{Methods}).
    In this setting, all methods perform uniformly worse than in the human genome simulations,
    with consistent underestimation of the mean selection coefficient
    and overestimation of the shape parameter (Figure \ref{fig:vaquita-dfe}A-B, Table \ref{tab:dfe_table}).
    Similar patterns are observed when re-running the simulations with a simpler demographic model
    without population size changes (Figure \ref{fig:vaquita-dfe.constant}, Table \ref{tab:dfe_table}).
    We suspect that this is due to the fact that all DFE-inference methods used here assume an additive model ($h = \frac{1}{2}$),
    while mutations in the simulated genomes were more recessive ($h < \frac{1}{2}$), and their dominance decreased with the strength of selection.
    To examine the influence of this model misspecification, we compared the inferred distributions of $s$ to the
    simulated distribution of $2hs$, which would have been equal to $s$ under an additive model (Figure \ref{fig:vaquita-dfe}C).
    Reassuringly, all three methods infer distributions that appear to fit the main mode of the simulated distribution of $2hs$.
    This is likely because deleterious alleles are typically at low frequency and thus heterozygous,
    where their selective effect is $h s$ rather than $s$.

@@

\section*{Discussion}
    \label{Discussion}
    In this paper, we have presented an important new addition to the \stdpopsim{} library
    that enables the simulation of genetic variation in the presence of selection.
    We have demonstrated the utility of this new API by showing how it can be used to benchmark
    the performance of different methods for demographic history inference, the inference of the distribution
    of fitness effects of new mutations, and the detection of selective sweeps.
    This allows for a wide range of models of selection to be easily simulated and compared to
    empirical data, with reproducibility, computational efficiency, and rigor.

@@

    Although we have shown how the new \stdpopsim API can be used to benchmark the performance of different methods,
    the benchmarks shown here merely scratch the surface of what can be done with the new API.
    We envision that, moving forward, many researchers will use
    the new API to simulate data under a wide range of models of selection, demography, and genome architecture, and to develop
    new methods for the analysis of genetic data in the presence of selection.
    The additions described here further expand the utility of \stdpopsim for both
    characterizing the patterns of diversity expected under a wide range of population genetic
    models, and for benchmarking methods for the analysis of genetic data
    on common ground in a way that was not possible before.

@@

    To facilitate direct comparison between inference methods,
    we scale each program's output to obtain a selection coefficient ($s$) consistent with the one used by \slim.
    \grapes outputs the mean and shape of the inferred distribution of $4N_es$.
    We thus scale the mean by a factor of  $\frac{\mu}{\theta}$,
    where $\mu$ is the assumed mutation rate and
    $\theta=4N_{e}\mu$ is the population-scaled mutation rate, as estimated by \grapes.
    \polydfe is similarly scaled by $\frac{2\mu}{\theta}$,
    reflecting the fact that it defines the selection coefficient such that a homozygote has fitness $1+ 2s$.
    \dadicli defines the selection coefficient similarly to \polydfe and infers a gamma distribution for $2 N_e s$, parameterized by the shape ($\alpha$) and rate ($\nu$).
    It also infers the total population-scaled mutation rate for selected sites over the considered region $\theta = 4 N_e \mu\, L\, p$,
    where $L$ is the total number of sites available for the analysis,
    and $p$ is the proportion of sites under selection.
    %IG: we're still missing justification for the scaling by L and p, but we can leave this for revision
    Thus, the mean value of $s$ in \slim terms is obtained by  $\frac{4\mu}{\theta} \alpha \nu L p$.
    Complete details for the DFE inference analysis can be found in the
    \snakemake workflow at \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/dfe.snake}.

>>>>

\section*{Abstract}
    \label{abstract}
    Selection is a fundamental evolutionary force that shapes patterns of genetic variation across species.
    However, simulations incorporating realistic selection along heterogeneous genomes in complex demographic histories
    are challenging, limiting our ability to benchmark statistical methods aimed at detecting selection and to explore theoretical predictions.
    \stdpopsim is a community-maintained simulation library that already provides an extensive catalog of
    species-specific population genetic models.
    Here we present a major extension to the \stdpopsim{} framework that enables simulation of various modes
    of selection, including background selection, selective sweeps,
    and arbitrary distributions of fitness effects (DFE) acting on annotated subsets of the genome (for instance, exons).
    This extension maintains \stdpopsim's core principles of reproducibility and accessibility while adding support
    for species-specific genomic annotations and published DFE estimates.
    We demonstrate the utility of this framework by comparing methods for demographic inference,
    DFE estimation, and selective sweep detection across several species and scenarios.
    Our results demonstrate the robustness of demographic inference methods to selection on linked sites,
    reveal the sensitivity of DFE-inference methods to model assumptions,
    and show how genomic features, like recombination rate and functional sequence density, influence power to detect selective sweeps.
    This extension to \stdpopsim{} provides a powerful new resource for the population genetics community
    to explore the interplay between selection and other evolutionary forces in a reproducible, user-friendly framework.

@@

    As with other aspects of \stdpopsim, we intend the selection models to be as biologically accurate as possible.
    Thus, the \stdpopsim catalog provides \texttt{Annotation} objects based on species' publicly available functional genomic elements
    and \texttt{DFE} objects based on published DFE estimates (Figure \ref{fig:schematic}A).
    By using annotation objects, one can, for example, simulate populations where selected mutations (whose selection
    coefficients are drawn from a specified DFE) occur within annotation regions (e.g. exons), while all mutations
    occurring outside of these regions do not affect fitness.
    There are relatively few species with published DFE estimates in the literature.
    \stdpopsim currently has implemented DFEs for five species in the catalog:
    \textit{Arabidopsis thaliana} \citep{huber2018gene},
    \textit{Drosophila melanogaster} \citep{ragsdale2016triallelic,huber2017determining,zhen2021greater},
    humans \citep{huber2017determining,kim2017inference,kyriazis2023using,castellano2019compmarison,castellano2019compmarison,rodrigues2024shared},
    \textit{Mus musculus} \citep{booker2021selective},
    and the vaquita porpoise \textit{Phocoena sinus} \citep{robinson2022critically}.
    While the DFEs are from a limited number of species, these DFEs can be applied to other species
    if and when it is suitable,
    namely, when a species of interest lacks a pubished DFE in the catalog \revpoint{1}{8}.
    For instance, one could simulate a butterfly species with a gamma-distributed DFE originally
    estimated from humans, or a human population with a DFE estimated from \textit{Drosophila}.
    Indeed, the DFE labeled \texttt{HomSap/Mixed\_K23} is recommended by \citet{kyriazis2023using}
    for ``general'' use across mammals, when lacking a more specific DFE.
    Furthermore, the user can specify a custom DFE and provide their own annotations
    of functional elements to simulate selection in a species for which we do not yet have
    a published DFE included in the catalog.

@@

    To expand these observations to additional species and to highlight the ease of comparison enabled by \stdpopsim,
    we performed a similar analysis using the vaquita porpoise \textit{Phocoena sinus}. The vaquita offers an intriguing contast to the demographic history of humans. While human poulations continue to expand, the vaquita is critically endangered, with only approxmiately 10 individauls remaining. Previous demogrpahic inference from genetic data \textcite{robinson2022critically} found a long-term effective popualtion size of 4,485 indiviuals which then decreased to 2,807 individuals 2,162 generations ago. The population then severley decreased even more to its current size in the last few generations. We simulated 100 genomes under a two-epoch demographic model inferred for vaquita by \textcite{robinson2022critically},
    and a constant recombination rate across the genome (see \textbf{Methods}).
    In simulations with selection, we applied the DFE inferred by \textcite{robinson2022critically} to exons
    annotated from the vaquita genome assembly.
    This DFE implements a relationship between selection and dominance
    in which very deleterious mutations (with selection coefficient $s<-0.1$) are fully recessive
    (dominance coefficient $h=0$) and weakly deleterious mutations ($s\geq -0.001$) are nearly
    additive ($h=0.4$; see \textbf{Methods}).
    As in the human simulations, we see that the presence of selection
    decreases polymorphism in genomes simulated with selection, but that the effect
    is only readily apparent within exonic regions (Figure \ref{fig:vaquita-summary-stats}).
    For these simulations, the SFS-based method (\stairway) produces more accurate
    inference than the two methods based on the sequential Markovian coalescent
    (\smcpp and \msmc; see Figure \ref{fig:vaquita-demography}).
    Regardless of these differences, we see that background selection again does not
    appear to considerably influence the inference of any of the four methods for this species,
    at least for the parameterization used here,
    even though vaquita has a very different evolutionary history than humans.

@@

    Next, we examined the performance of these three DFE-inference methods when assuming a different DFE. To do this, we again turned to the vaquita. Unlike the human simulations where all mutations were assumed to be additive (h=0.5), the vaquita DFE applied a gamma-distributed selection coefficient to mutations in exons
    with a relationship between the selection and dominance coefficients (see \textbf{Methods}). Specifically, more deleterious mutations in the vaquita DFE are assumed to be more recessive. 
    In this setting, all methods perform uniformly worse than in the human genome simulations,
    with consistent underestimation of the mean selection coefficient
    and overestimation of the shape parameter (Figure \ref{fig:vaquita-dfe}A-B, Table \ref{tab:dfe_table}).
    Similar patterns are observed when re-running the simulations with a simpler demographic model
    without population size changes (Figure \ref{fig:vaquita-dfe.constant}, Table \ref{tab:dfe_table}).
    We suspect that this is due to the fact that all DFE-inference methods used here assume an additive model ($h = \frac{1}{2}$),
    while mutations in the simulated genomes were more recessive ($h < \frac{1}{2}$), and their dominance decreased with the strength of selection.
    To examine the influence of this model misspecification, we compared the inferred distributions of $s$ to the
    simulated distribution of $2hs$, which would have been equal to $s$ under an additive model (Figure \ref{fig:vaquita-dfe}C).
    Reassuringly, all three methods infer distributions that appear to fit the main mode of the simulated distribution of $2hs$.
    This is likely because deleterious alleles are typically at low frequency and thus heterozygous,
    where their selective effect is $h s$ rather than $s$.

@@

\section*{Discussion}
    \label{Discussion}
    In this paper, we have presented an important new addition to the \stdpopsim{} library
    that enables the simulation of genetic variation in the presence of selection.
    We have demonstrated the utility of this new API by showing several illustritive exampls of how it can be used to benchmark
    the performance of different methods for demographic history inference, the inference of the distribution
    of fitness effects of new mutations, and the detection of selective sweeps.
    This allows for a wide range of models of selection to be easily simulated and compared to
    empirical data, with reproducibility, computational efficiency, and rigor.

@@

    Although we have shown how the new \stdpopsim API can be used to benchmark the performance of different methods,
    the comparisons shown here merely scratch the surface of what can be done with the new API.
    We envision that, moving forward, many researchers will use
    the new API to simulate data under a wide range of models of selection, demography, and genome architecture, and to develop
    new methods for the analysis of genetic data in the presence of selection.
    The additions described here further expand the utility of \stdpopsim for both
    characterizing the patterns of diversity expected under a wide range of population genetic
    models, and for benchmarking methods for the analysis of genetic data
    on common ground in a way that was not possible before.

@@

    To facilitate direct comparison between inference methods,
    we scale each program's output to obtain a selection coefficient ($s$) consistent with the one used by \slim.
    \grapes outputs the mean and shape of the inferred distribution of $4N_es$.
    We thus scale the mean by a factor of  $\frac{\mu}{\theta}$,
    where $\mu$ is the assumed mutation rate and
    $\theta=4N_{e}\mu$ is the population-scaled mutation rate, as estimated by \grapes.
    \polydfe is similarly scaled by $\frac{2\mu}{\theta}$,
    reflecting the fact that it defines the selection coefficient such that a homozygote has fitness $1+ 2s$.
    \dadicli defines the selection coefficient similarly to \polydfe and infers a gamma distribution for $2 N_e s$, parameterized by the shape ($\alpha$) and rate ($\nu$).
    It also infers the total population-scaled mutation rate for selected sites over the considered region $\theta = 4 N_e \mu\, L\, p$,
    where $L$ is the total number of sites available for the analysis,
    and $p$ is the proportion of sites under selection.
    Thus, the mean value of $s$ in \slim terms is obtained by  $\frac{4\mu}{\theta} \alpha \nu L p$.
    Complete details for the DFE inference analysis can be found in the
    \snakemake workflow at \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/dfe.snake}.
